---
title: "Detect/check for separation and infinite maximum likelihood estimates in logistic regression"
author: "[Ioannis Kosmidis](http://www.ikosmidis.com) and [Dirk Schumacher](http://www.dirk-schumacher.net)"
date: "5 January 2020"
output: rmarkdown::html_vignette
bibliography: detectseparation.bib
vignette: >
  %\VignetteIndexEntry{Detecting separation and infinite estimates in logistic regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6
)
```


# The **detectseparation** package

[**detectseparation**](https://github.com/ikosmidis/detectseparation)
provides *pre-fit* and *post-fit* methods for the detection of
separation and of infinite maximum likelihood estimates in binomial
response generalized linear models.

The key methods are `detect_separation` and `check_infinite_estimates`
and this vignettes describes their use.

# Checking for infinite estimates

@heinze+schemper:2002 used a logistic regression model to analyze data
from a study on endometrial cancer [see, @agresti:2015, Section 5.7 or
`?endometrial` for more details on the data set]. Below, we refit the
model in @heinze+schemper:2002 in order to demonstrate the
functionality that **detectseparation** provides.

```{r, echo = TRUE, eval = TRUE}
library("detectseparation")
data("endometrial", package = "detectseparation")
endo_glm <- glm(HG ~ NV + PI + EH, family = binomial(), data = endometrial)
theta_mle <- coef(endo_glm)
summary(endo_glm)
```

The maximum likelihood (ML) estimate of the parameter for `NV` is actually
infinite. The reported, apparently finite value is merely due to false
convergence of the iterative estimation procedure. The same is true
for the estimated standard error, and, hence the value `r
round(coef(summary(endo_glm))["NV", "z value"], 3)` for the $z$-statistic
cannot be trusted for inference on the size of the effect for `NV`.

@lesaffre+albert:1989[, Section 4] describe a procedure that can hint
on the occurrence of infinite estimates. In particular, the model is
successively refitted, by increasing the maximum number of allowed
iteratively re-weighted least squares iterations at east step. The
estimated asymptotic standard errors from each step are, then, divided
to the corresponding ones from the first fit. If the sequence of
ratios diverges, then the maximum likelihood estimate of the
corresponding parameter is minus or plus infinity. The following code
chunk applies this process to `endo_glm`.

```{r, echo = TRUE, eval = TRUE}
(inf_check <- check_infinite_estimates(endo_glm))
plot(inf_check)
```

Clearly, the ratios of estimated standard errors diverge for `NV`.

# Detecting separation

`detect_separation` tests for the occurrence of complete or
quasi-complete separation in datasets for binomial response
generalized linear models, and finds which of the parameters will have
infinite maximum likelihood estimates. `detect_separation` relies on
the linear programming methods developed in the 2017 PhD thesis by
Kjell Konis [@konis:2007].

`detect_separation` is *pre-fit* method, in the sense that it does not
need to estimate the model to detect separation and/or identify
infinite estimates. For example

```{r, eval = TRUE, echo = TRUE}
endo_sep <- glm(HG ~ NV + PI + EH, data = endometrial,
                family = binomial("logit"),
                method = "detect_separation")
endo_sep
```

The `detect_separation` method reports that there is separation in the
data, that the estimates for `(Intercept)`, `PI` and `EH` are finite
(coded 0), and that the estimate for `NV` is plus infinity. So, the
actual maximum likelihood estimates are

```{r, echo = TRUE,  eval = TRUE}
coef(endo_glm) + coef(endo_sep)
```

and the estimated standard errors are

```{r, echo = TRUE,  eval = TRUE}
coef(summary(endo_glm))[, "Std. Error"] + abs(coef(endo_sep))
```

We can also use the
[`glpk`](https://CRAN.R-project.org/package=ROI.plugin.glpk) solver
for solving the linear program for separation detection

```{r, echo = TRUE, eval = TRUE}
update(endo_sep, solver = "glpk")
```

or use the implementation using
[**lpSolveAPI**](https://CRAN.R-project.org/package=ROI.plugin.glpk)
directly

```{r,  echo = TRUE, eval = TRUE}
update(endo_sep, implementation = "lpSolveAPI")
```
See `?detect_separation_control` for more options.

As proven in [@kosmidis+firth:2018], an estimator that is always
finite, regardless whether separation occurs on not, is the
reduced-bias estimator of [@firth:1993], which is implemented in the
[**brglm2**](https://CRAN.R-project.org/package=brglm2) R package.

```{r, echo = TRUE, eval = TRUE}
library("brglm2")
summary(update(endo_glm, method = "brglm_fit"))
```

# Citation
If you found this vignette or **detectseparation** useful, please
consider citing **detectseparation**. You can find information on how
to do this by typing `citation("detectseparation")`.

# References
